Calibrating seed-based alignment heuristics with Sesame

### 2020-02-26 manuscript.tex
Add appendix with proofs

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Figure~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location using MEM seeds (even though it is the best
location).

@@

\begin{definition}
\label{def:seg}
Every symbol of the alphabet that is different from $\square$ is called a
terminator. A segment is a terminator preceded by 0 or more symbols $\square$.
The last segment of the read is terminated by the symbol $|$ and is called the
tail.
\end{definition}

Since the decomposition of the read in segments is unique, we can view a
read as a sequence of segments with a tail, instead of a sequence of
nucleotides. Figure~\ref{fig:simple} shows an example of decomposition in
segments. On-target seeds cannot contain sequencing errors, therefore they
must be completely embedded inside a segment. So the sizes of the segments
indicate whether the read contains an on-target seed or not.

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) and the probability of occurrence of the symbol
$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that the weighted generating function of a segment
with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
the weighted generating function of a tail segment with $i$ symbols
$\square$ is $(qz)^i$.

The key insight is that reads without on-target seed are exactly the reads
that are made of segments with fewer than $\gamma$ symbols $\square$. The
weighted generating function of such segments is $pz \cdot \big(
1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
$1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
encoded in a transfer matrix.

There are two kinds of objects: the $\Downarrow$ segments and the tails,
so the dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$
segment can be followed by another $\Downarrow$ segment or by the tail.
The tail cannot be followed by anything. The expression of the transfer
matrix $M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

In the representation above, the different types of objects are identified
by their terminator. The terminators are written in the margins of the
transfer matrix for readability.

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term --- associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by $\Downarrow$ segments and only
by those (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. It is easy to check that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
\end{equation}

This function can be expanded as a Taylor series $a_0 + a_1z + a_2z^2 +
\ldots$, but the coefficients $a_0, a_1, a_2, \ldots$ are unknown. By
construction, $a_k$ is the quantity of interest, \textit{i.e.} it is the
probability that a read of size $k$ does not contain an on-target seed, so
we now need to extract the coefficients from expression (\ref{eq:F}).
There are several methods to do so; the one we choose here is to build a
recurrence equation to compute $a_k$ from the previous terms of the series
$a_0, a_1, \ldots, a_{k-1}$. For this, we observe that for $z \neq 1/q$,
equation
\begin{equation*}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} =
  a_0 + a_1z +a_2z^2 + \ldots
\end{equation*}
is equivalent to
\begin{equation*}
(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
\end{equation*}
Balancing the terms of same degree on both sides of the equation leads
to the following relationship between the coefficients of the series:
\begin{equation}
\label{eq:recur}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -q^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}

These results are consistent: for $k < \gamma$, the read is too short so
the probability that it contains no seed is 1; for $k = \gamma$, the
read contains a seed if and only if it has no error, which occurs with
probability $q^\gamma$.

@@

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Prob. seeding error & Prob. other error & \% seeding \\
\hline
50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
\end{tabular}
\caption{My favorite table. $p = 0.01$ and $\mu = 0.06$.}
\label{table_mem}
\end{table}

@@

Tables~\ref{table_exact} and \ref{table_skip} shows the result of numeric
simulations for exact and skip-9 seeds, respectively. It appears that in
these cases, the probability that probability of mapping the read to an
incorrect location (\textit{i.e.} not the true location) is substantially
higher than the probability that seeding is off target. As a consequence,
seeding probabilities are not very useful to address true mapping problem
when exact and skip seeds are used. In this case, it is more accurate to
use the probabilities that the true location is not the best.

@@

>>>>

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the minimum seed size $\gamma$.
Figure~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location using MEM seeds (even though it is the best
location).

@@

\begin{definition}
\label{def:seg}
A terminator is any symbol that is different from the symbol $\square$. A
segment is a sequence of 0 or more symbols $\square$ followed by a
terminator. The tail is the last segment of the read, where the terminator
is always the special symbol $|$.
\end{definition}

Since the decomposition of the read in segments is unique, we can view a
read as a sequence of segments with a tail, instead of a sequence of
nucleotides. Figure~\ref{fig:simple} shows an example of decomposition in
segments. On-target seeds cannot contain sequencing errors, therefore they
must be completely embedded in a segment. So the sizes of the segments
indicate whether the read contains an on-target seed or not.

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) so the probability of occurrence of the symbol
$\square$ is $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that a segment of $i$ symbols $\square$ followed by
a terminator has weighted generating function $(qz)^ipz$. The final symbol
$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
symbol $|$ has weighted generating function $(qz)^i$.

The key insight is that the reads without on-target seed are exactly the
reads that are made of segments with fewer than $\gamma$ symbols
$\square$, where $\gamma$ is the minimum seed size. The weighted
generating function of such segments is $\big( 1+qz+\ldots+(qz)^{\gamma-1}
\big) pz$, and that of the tail is $1+qz+\ldots+(qz)^{\gamma-1}$. This
gives a construction plan that can be encoded in a transfer matrix.

Reads consist of only two kinds of objects: the segments terminated by
$\Downarrow$ (or $\Downarrow$-segments for short) and the tails, so the
dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$-segment
can be followed by another $\Downarrow$-segment or by the tail. The tail
cannot be followed by anything. The expression of the transfer matrix
$M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

In the representation above, the different types of segments are
identified by their terminator. The terminators are written in the margins
of the transfer matrix for readability.

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term---associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by $\Downarrow$-segments and only
by those (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. In appendix~\ref{app_ex1}, we show that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz}.
\end{equation}

This function can be expanded as a Taylor series $a_0 + a_1z + a_2z^2 +
\ldots$, but the coefficients $a_0, a_1, a_2, \ldots$ are unknown. By
construction, $a_k$ is the quantity of interest, \textit{i.e.} it is the
probability that a read of size $k$ does not contain an on-target seed of
minimum size $\gamma$, so we now need to extract the Taylor coefficients
from expression (\ref{eq:F}). There are several methods to do so; the one
we choose here is to build a recurrence equation. In
appendix~\ref{app_rec}, we show that

\begin{equation}
\label{eq:recur}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -q^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}

Note that for $k < \gamma$, the read is too short so the probability that
it contains no seed is 1; for $k = \gamma$, the read contains a seed if
and only if it has no error, which occurs with probability $q^\gamma$.

@@

\begin{table}
% MEM seeds.
\renewcommand{\arraystretch}{1.1}
\centering
\begin{tabular}{ccccc}
k & N & Not seeded & Seeded, not best & \% seed failure \\
\hline
50  &  1  & $7.8 \cdot 10^{-4}$ & $9.5 \cdot 10^{-5}$ & $89.1\%$ \\
50  & 10  & $7.8 \cdot 10^{-3}$ & $9.1 \cdot 10^{-4}$ & $89.6\%$ \\
50  & 100 & $7.6 \cdot 10^{-2}$ & $5.7 \cdot 10^{-3}$ & $93.0\%$ \\
100 &  1  & $2.1 \cdot 10^{-5}$ & $7.6 \cdot 10^{-6}$ & $73.7\%$ \\
100 & 10  & $2.6 \cdot 10^{-4}$ & $7.1 \cdot 10^{-5}$ & $78.1\%$ \\
100 & 100 & $5.3 \cdot 10^{-3}$ & $5.9 \cdot 10^{-4}$ & $90.1\%$
\end{tabular}
\caption{Numerical simulating of mapping errors with MEM seeds. Reads are
mapped incorrectly if either the true location is not in the candidate set
(not seeded) or it is in the candidate set but is not the best location
(seeded, not best). Ideal reads were randomly generated 1,000,000 times
with sequencing error rate $p = 0.01$, divergence rate $\mu = 0.06$,
varying the read size $k$ and number $N$ of duplicates as indicated in the
table. Mapping errors of both types were recorded, and the fraction of the
first to the total was computed.}
\label{table_mem}
\end{table}

@@

Tables~\ref{table_exact} and \ref{table_skip} shows the result of numeric
simulations for exact and skip-9 seeds, respectively. It appears that in
these cases, the probability that probability of mapping the read to an
incorrect location (\textit{i.e.} not the true location) is substantially
higher than the probability that seeding is off target. As a consequence,
seeding probabilities are not very useful to address true mapping problem
when exact and skip seeds are used. In this case, it is more accurate to
use the probabilities that the true location is not the best, because
they are in general much higher.

In conclusion, the true location problem can be calibrated using the
methods presented here when using MEM seeds, and by computing the
probability that the true location is not the best when using exact or
skip-seeds.

@@

\appendix
\section{Appendix}
\subsection{Derivation of expression~(\ref{eq:F})}
\label{app_ex1}

In section~\ref{sec:example_exact}, we had to compute the top right entry
of the matrix $M_*(z)$ defined as $M_*(z) = M_0(z) + M_0(z)^2 + \ldots =
M_0(z) \cdot (I-M_0(z))^{-1}$, where
\begin{equation*}
M_0(z) =
\begin{bmatrix}
\big( 1+qz+\ldots+(qz)^{\gamma-1}
\big) pz  & 1+qz+\ldots+(qz)^{\gamma-1} \\
0 & 0 \\
\end{bmatrix}.
\end{equation*}

Since $I-M_0(z)$ is a $2\times2$ matrix, we can invert it analytically.
Observing that the determinant of $I-M_0(z)$ is $\Delta(z) = 1-\big(
1+qz+\ldots+(qz)^{\gamma-1}\big) pz$, we find
\begin{equation*}
(I-M_0(z))^{-1} = \frac{1}{\Delta(z)}
\begin{bmatrix}
1 & 1+qz+\ldots+(qz)^{\gamma-1} \\
0 & 1-\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz
\end{bmatrix}.
\end{equation*}

Premultiplying by $M_0(z)$, we finally obtain
\begin{equation*}
M_*(z) = \frac{1}{\Delta(z)}
\begin{bmatrix}
\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz &
1+qz+\ldots+(qz)^{\gamma-1} \\
0 & 0 \end{bmatrix},
\end{equation*}
where the top right term is equal to
\begin{equation}
\tag{\ref{eq:F}}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
{1-\big( 1+qz+\ldots+(qz)^{\gamma-1} \big) pz}.
\end{equation}

\subsection{Derivation of expression~(\ref{eq:recur})}
\label{app_rec}

In section~\ref{sec:example_exact}, we had to extract the Taylor
coefficients $a_0, a_1, a_2 \ldots$ of the function
\begin{equation*}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz} =
  a_0 + a_1z +a_2z^2 + \ldots
\end{equation*}

Multipying both terms of the equation by $1-pz
\big(1+qz+\ldots+(qz)^{\gamma-1} \big)$ we obtain
\begin{equation*}
1+qz+\ldots+(qz)^{\gamma-1} =
\left(1-\big(1+qz+\ldots+(qz)^{\gamma-1} \big) pz\right)
(a_0 + a_1z +a_2z^2 + \ldots).
\end{equation*}

Multiplying again both terms of the equation by $1-qz$, we find
\begin{eqnarray*}
1-(qz)^\gamma &=& (1-qz) \left(1- \big(1+qz+\ldots+(qz)^{\gamma-1}
\big) pz\right) (a_0 + a_1z +a_2z^2 + \ldots) \\
&=& (1-z+pq^\gamma \cdot z^{\gamma+1}) (a_0 + a_1z +a_2z^2 + \ldots) \\
&=& a_0 + \sum_{k=1}^{\gamma} (a_k-a_{k-1})z^k +
  \sum_{k=\gamma+1}^{\infty} (a_k-a_{k-1}+ pq^\gamma \cdot
a_{k-\gamma-1})z^k.
\end{eqnarray*}

Balancing the terms of degree 0 on both sides of the equation forces $a_0
= 1$. The terms of degree 1 to degree $\gamma-1$ are null on the left
side, which implies $a_k - a_{k-1} = 0$ and consequently $a_k = a_0 = 1$
for $k = 1, 2, \ldots, \gamma-1$. The term of degree $\gamma$ is
$-q^\gamma$ on the left side, so $a_\gamma-a_{\gamma-1} = -q^\gamma$, or
$a_\gamma = 1-q^\gamma$. Finally, all the other terms on the left side
are null, which implies $a_k-a_{k-1}+ pq^\gamma \cdot a_{k-\gamma-1} = 0$,
or $a_k = a_{k-1}-pq^\gamma\cdot a_{k-\gamma-1}$ for $k = \gamma+1,
\gamma+2 \ldots$

In comclusion, the Taylor coefficients $a_0, a_1, a_2, \ldots$ satisfy the
recurrence equation
\begin{equation}
\tag{\ref{eq:recur}}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -q^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}
